\thetitle

### 2020-11-20 writing/rational.tex
reivising rational and sample for the new stats

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails if this component causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenarios successfully pinpoints the location of
bugs. 
\begin{quote}
\it A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff its last scenario $\conf_n$ does not type check.  A natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is failing natural blame trails are those that have not reached a scenario that
reveals the bug at compile time and their last scenario does not produce blame. Thus
the rational programmer receives no hints from the gradual type system on how to
continue the search for the bug. In contrast, if the rational programmer
knows that for a program  $\system$
the lattice $\lattice{\system}$ contains only successful natural blame
trails, then the rational programmer can rely on natural blame while debugging
a type-level mistakes in $\system$ independently of $\system$'s type
annotations.

While the absence of failing natural blame trails in $\lattice{\system}$
imply that
pays off to heed blame assignments during debugging $\system$, it does not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing the top of stack trace of a failed run time type check, dubbed the
location of the exception, might be as useful as typing the blamed one.

@@

Putting the two modes together, we can now compare the usefulness of blame 
and mere exceptions.
\begin{quote}
\it 
  Given a program $\system$, natural exceptions are  \emph{at least as useful} as 
  natural blame for $\system$ iff  for every interesting scenario 
  $\conf_0$ in $\lattice{\system}$ when the natural blame trail 
  that starts at $\conf_0$ is successful then the natural exception trail that
  starts at $\conf_0$ is also successful.
\end{quote}
\indent Of course if natural exceptions are not at least as useful as
natural blame for $\system$ then we have evidence that blame 
is \emph{valuable} for the rational programmer in the context of natural.

@@

The definition of \emph{transient exception trails} 
is analogous to that for natural and, same as for natural, we use it to  
specify when transient exceptions are at least as useful as the 
transient-last and transient-first blame.

@@

Trails and their properties give us the tools for a rigorous examination
of blame for natural, transient and erasure Typed Racket in the setting of the 
mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
our experimental process collects data to answer three initial questions for each mutant:
\begin{itemize}
\item[$Q_1$] Is blame valuable in the context of natural?

\item[$Q_2$] Is first blame valuable in the context of transient?

\item[$Q_3$] Is last blame valuable in the context of transient?

@@

Furthermore, our experiment can compare the relative usefulness of blame
between any two of the three semantics. 
\begin{itemize}
\item[$Q_*$] Is blame for X at least as useful as blame for Y where X, Y is natural,
  erasure or natural?
\end{itemize}

@@

In detail we answer $Q_1$ for an ill-typed mutant from
section~\ref{sec:mutate} by analyzing the natural blame and
natural exception trails in the scenario lattice of the mutant. A
log reports on success and failure endings.  The first step is to construct the
mutant's scenario lattice and select one of the interesting debugging
scenarios from section~\ref{sec:mutate}.
Our test
bed extends these trails according to the natural-blame programmer.  If no scenarios
can be added to the trail, we check if
the last scenario of the trail type checks or not. If it does, this natural-blame
trail is successful, otherwise it is failing. We repeat the same process for the
same roots once more following the natural-exception programmer. 
Next, based on the collected data, we calculate the
success/failure results for the trails to determine whether exceptions are
as useful as blame for this mutant. Finally, we
obtain a positive answer for
$Q_1$ when there are mutants where natural exceptions are not as useful 
as natural blame.  The process is analogous for $Q_2$ and $Q_3$  employing the respective
modes of the rational programmer. For $Q_*$, we also proceed in a similar
manner except that we use modes 
of the rational programmer from different semantics. For instance to
decide whether blame for natural is as useful as blame for
transient, we compare the success of
natural blame trails with that of the transient first and transient last
blame trails and we determine if one mode is as useful as another. 
Figure~\ref{fig:process} summarizes the
experimental process for one mode of the rational programmer and connects
it with the mutations from section~\ref{sec:mutate}.

@@

A comparison of the effort of different modes of the rational programmer
can help shed some further light to the effectiveness of the
three gradual typing systems. For example, consider one of 
our mutants.  If, for instance, both natural blame and transient first
blame are equally useful for the mutant, the two nodes of the 
rational programmer can compete to see which one
debugs the interesting scenarios of the mutant with less effort. In
general, if the effort distribution of a mode of the rational programmer for a
mutant has a shorter tail and more
volume around smaller values compared to the effort distribution of another
mode, then the first mode is the more effective of the two.

Finally, effort allows us to validate that the observed effectiveness 
of the rational programmer is not an artifact of pure chance.
To test the null hypothesis for the effectiveness of the rational
programmer, we can simply compare the effort distribution of a mode of the
rational programmer with a rational programmer that ignores blame and
exceptions and instead selects which component to type next randomly. 
Of course, since each mutant has a finite number of components, 
such a random mode is always useful for all mutants in the context of all
three semantics. At the same time though,
the natural, transient and erasure random mode exhibit a normal effort distribution for each mutant
that should be distinct from the effort distribution of all other modes of
the rational programmer if the effectiveness of the latter is not
coincidental.

>>>>

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails if this component causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenarios successfully pinpoints the location of
bugs. 
\begin{quote}
\it A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff its last scenario $\conf_n$ does not type check.  A natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is failing natural blame trails are those that have not reached a scenario that
reveals the bug at compile time and their last scenario does not produce blame. Thus
the rational programmer receives no hints from the gradual type system on how to
continue the search for the bug.

While a successful natural blame trail indicates that it 
pays off to heed blame assignments during debugging the trail's root, it does not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing the top of stack trace of a failed run time type check, dubbed the
location of the exception, might be as useful as typing the blamed one.

@@

Putting the two modes together, we can now compare the usefulness of blame 
and mere exceptions for debugging a scenario in the context of natural semantics.
\begin{quote}
\it 
  Given a program $\system$ and a root $\conf_0$ in $\lattice{\system}$,
  natural blame is \emph{more useful} than natural exceptions for
  debugging $\conf_0$ iff 
  the natural blame trail 
  that starts at $\conf_0$ is successful while the natural exception trail that
  starts at $\conf_0$ is failing.
\end{quote}

@@

The definition of \emph{transient exception trails} 
is analogous to that for natural and, same as for natural, we use it to  
specify when  
transient-last and transient-first blame are more useful than 
transient exceptions for debugging a scenario.

@@

Trails and their properties give us the tools for a rigorous examination
of blame for natural, transient and erasure Typed Racket in the setting of the 
mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
our experimental process collects data to answer three initial questions
for our interesting debugging scenarios:
\begin{itemize}
\item[$Q_1$] Is blame useful in the context of natural?

\item[$Q_2$] Is first blame useful in the context of transient?

\item[$Q_3$] Is last blame useful in the context of transient?

@@

Furthermore, our experiment can compare the relative usefulness of blame
between any two of the three semantics. 
\begin{itemize}
\item[$Q_*$] Is blame for X more useful than blame for Y where X, Y is natural,
  erasure or natural?
\end{itemize}

@@

In detail we answer $Q_1$  by comparing the success of the 
natural blame and natural exception trails for all interesting 
debugging scenarios from section~\ref{sec:mutate}.
  The first step is to construct each
mutant's scenario lattice and identify their interesting debugging
scenarios.
Our test
bed extends the trails that start from such roots according to the natural-blame programmer.  If no scenarios
can be added to the trail, we check if
the last scenario of the trail type checks or not. If it does, we log natural-blame
trail is successful, otherwise that is failing. We repeat the process for the
same roots again but this time for the other mode, natural exceptions.
 Figure~\ref{fig:process}
summarizes this
experimental process for one mode of the rational programmer and connects
it with the mutations from section~\ref{sec:mutate}.
Next, based on the collected data, we calculate the
success/failure results for the trails to determine for each root whether natural blame
is more useful than natural exceptions. We
obtain a positive answer for
$Q_1$ when a root exists where the above is true since it is evidence that
there is at least one interesting scenario that the rational programmer
manages to debug because of blame. 
The process is analogous for $Q_2$ and $Q_3$  employing the respective
modes of the rational programmer.

For $Q_*$, the process is a bit more
involved. At a first level we proceed in a similar
manner except that we use two modes 
of the rational programmer from different semantics.
However to answer this question we compare the percentage of scenarios
where one mode is more useful than the other and the inverse.
For instance to
decide whether blame for natural is more useful than first blame for transient, 
we compare the percentage of interesting scenarios where 
natural blame is more successful than transient first blame 
with the percentage of interesting scenarios where transient first blame 
is more successful than natural blame. 
Finally we repeat these steps one more time to compare natural blame and transient last blame
and get a complete picture of the comparative usefulness of blame
in the two semantics.

@@

A comparison of the effort of different modes of the rational programmer
can help shed some further light to the effectiveness of the
three gradual typing systems. For example, consider one debugging scenario.  
If, for instance, both  natural blame and transient first
blame trails are successful, the two modes of the 
rational programmer can compete to see which one
debugs the interesting scenario with less effort. In
general, if the effort distribution of a mode of the rational programmer
has a shorter tail and more
volume around smaller values compared to the effort distribution of another
mode, then this is evidence that the first mode is the more effective of the two.

Finally, effort allows us to test whether the observed effectiveness 
of the rational programmer is an artifact of pure chance or not.
For that, we can simply compare the effort distribution of a mode of the
rational programmer with a rational programmer that ignores blame and
exceptions and instead selects which component to type next randomly. 
Of course, since each mutant has a finite number of components, 
such the random mode trails are always successful. At the same time though,
the natural, transient and erasure random mode exhibit a normal effort
distribution across scenarios that should be distinct from the effort distribution of all other modes of
the rational programmer if the effectiveness of the latter is not
coincidental.

### 2020-11-20 writing/sample.tex
reivising rational and sample for the new stats

The mutators for section~\ref{sec:mutate} generate 96,515,744 debugging scenarios; 
far too many to even identify the interesting ones among
them. Hence, we perform stratified uniform random sampling with two
levels of stratification to collect sufficiently many interesting
debugging scenarios for drawing statistically sound conclusions. 
The outcome of our sampling process is that we test each mode of the
rational programmer on 72,192 randomly selected interesting debugging
scenarios in total spread across both benchmarks and mutators.

As outlined in section~\ref{sec:rational}, our experimental questions
boil down whether a mode of the rational programmer is
as useful as another mode for all the interesting scenarios of a mutant $\system$.
Given the binary nature of this basic question, we can rely on random
sampling of interesting scenarios from $\lattice{\system}$ to obtain an
approximate answer.  Thus $\lattice{\system}$ becomes the first stratum
for our sampling; with a sufficiently large sample, we can obtain from the
sample an estimate of the usefulness of a mode for the entire lattice.
The estimate tells us with some confidence whether the mode is useful for all trails in the lattice (if all of the samples yield positive results) or none of them.
In fact, the estimate can say neither if some of our samples yield positive results and others yield negative ones; in that case we can say no more than that the mode is sometimes useful.
We select a sample size of 96 to obtain 95\% confidence and a 5\% margin of error for our estimates
(using standard statistical methods, the choice does not depend on the population size for sufficiently large populations).
For lattices that have less than 96 interesting scenarios, we select them all.

The second sampling stratum is the set of mutants. After all, our mutators
generate 16,800 mutants with interesting scenarios; still far too many to
explore even after sampling the lattice of each mutant. Thus we randomly
select from the mutants of each benchmark to obtain the largest
sample that keeps the experiment computationally feasible.
Our selection of mutants ensures that, per benchmark, the sample reflects the diversity of mutants with
respect to the mutators that generated them.  Specifically,  we sample 80
mutants with interesting scenarios per benchmark,  evenly-distributed
across all of the mutators that contribute mutants for the benchmark.

>>>>

The mutators from section~\ref{sec:mutate} generate 96,515,744 debugging
scenarios; far too many to even identify the interesting ones among them.
Furthermore this population is heterogeneous; different scenarios come
from different mutants and different mutants are the result of a
different mutator and benchmark pair. Hence, to make the experiment
computationally feasible and statistically precise, we perform stratified
uniform random sampling with two levels of stratification.  The first
stratum is the set of mutants and we randomly select from the mutants of
each benchmark.  Our selection of mutants ensures that, per
benchmark, the sample reflects the diversity of mutants with respect to
the mutators that generated them.  Specifically,  we sample 80 mutants
with interesting scenarios per benchmark,  evenly-distributed across all
of the mutators that contribute mutants for the benchmark.  The second
stratum of our sampling is the lattice of a mutant and we draw 96 random
interesting scenarios from each lattice. For lattices that have less than
96 interesting scenarios, we select them all.

In summary, our sampling process takes advantage of ancillary information
to deliver a representative sufficiently large set of scenarios. As a result we test each
mode of the rational programmer on 72,192 randomly selected interesting
debugging scenarios spread across both benchmarks and mutators.
Furthermore stratified random sampling allows us to derive conclusions
about the proportions of the whole population from the proportions of
our sample, exactly what we need to project comparisons about the
usefulness of various modes of the
rational programmer from the sample to set of all interesting debugging
scenarios.  Finally, we can also provide separate estimates per stratum and provide different
viewpoints of our results. For instance, because of the size of the sample
of scenarios from the lattice of a mutant, we can estimate whether natural
blame is more useful than natural exceptions for all interesting scenarios
of each mutant separately with 95\% confidence and 5\% error margin.
